---
title: "micro-RNA Heart Transplantation"
output:
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

### Data importing ###

```{r}
source(here("config.R"))
source(here("Transplantation", "config_t.R"))
source(here("src", "data_processing.R"))
source(here("src", "visualization.R"))
source(here("src", "calculations.R"))
data <- load_and_process_data()
counts <- data$counts
phenotable <- data$phenotable
summed_anno <- load_sample_annotation(phenotable = data$phenotable)
```
**Venn plot (>150 counts)**

```{r}
venn_plot <- create_venn_diagram(
  counts = data$counts,
  phenotable = data$phenotable
)
venn_plot$plot
```
**Unique miRs**
```{r}
get_unique_venn_elements(venn_plot$venn_list)
```
**Barplots**
```{r}
create_barplot_all_dataset(data = summed_anno)
```
```{r}
create_barplot_all_dataset_normalized(data = summed_anno)
```
```{r}
grouped_barplots <- create_grouped_barplots(data = summed_anno, phenotable = data$phenotable)
grouped_barplots$counts
#grouped_barplots$proportion
```
###  Run Differential Expression Analysis ###
```{r}
dds <- create_dds(counts, coldata = phenotable, ref_level = "NR", condition_col = "condition", check_rank = TRUE)
dds <- DESeq(dds, fitType = "parametric")
dds 
```
**transformation**

• The plot should show a nearly horizontal line, meaning the SD is independent of the mean expression.
• This means the transformation has successfully stabilized the variance, making the data more suitable for further analysis.
```{r}
rlt_list <- auto_transform_dds(dds)
rlt <- rlt_list$transformed
rlt_list
```
**PCA plot**
```{r}
create_pca_plot(rlt, condition_col = "condition")
```

**Plot a heatmap of most variable genes**
```{r}
create_heatmap_most_expressed(rlt, top_n = 30, group_col = "condition", output_file = NULL)
```
**Plot of the distance between samples heatmap**

```{r}
create_sample_distance_heatmap(rlt, group_col = "condition", output_file = NULL)
```
## Results ##
```{r}
all_res <- list()
all_sign_res <- list()
for (name in names(CONTRASTS)) {
  res <- get_significant_results(
    dds = dds,
    contrast = CONTRASTS[[name]],
    output_res = file.path("results", paste0("results_", name, ".csv")),
    output_sign_res = file.path("results", paste0("sign_results_", name, ".csv")),
    alpha = 0.05, lfc_threshold = 1)
  all_res[[name]] <- res$res_df         
  all_sign_res[[name]] <- res$res_sign_df
}

all_sign_res$humoral
all_sign_res$cellular
all_sign_res$CAV
```
**MA plot**
```{r}
ma_plots <- list()
for(name in names(all_res)) {
  ma_plots[[name]] <- plot_ma_from_res(
    res_df = all_res[[name]], 
    alpha = 0.05, 
    contrast_name = name,
    output_file = file.path("figures", paste0("ma_plot_", name, ".png"))
  )
}

ma_plots$humoral
ma_plots$cellular
ma_plots$CAV
```
**Volcano plot**

```{r}
volcano_plots <- list()
for(name in names(all_res)) {
  volcano_plots[[name]] <- create_volcano_plot(
    res_df = all_res[[name]],
    res_sign_df = all_sign_res[[name]],
    contrast_name = name,
    output_file = file.path("figures", paste0("volcano_", name, ".png"))
  )
}

volcano_plots$humoral
volcano_plots$cellular
volcano_plots$CAV
```

**Create boxplot for specific gene**

```{r}
create_gene_boxplot(dds, "hsa-miR-100-5p", output_file = NULL)
```
**Plot a heatmap of diff expressed genes**
```{r}
source(here("src", "visualization.R"))
heatmap_plots <- list()
for(name in names(all_res)) {
  volcano_plots[[name]] <- create_heatmap_diff_expressed(
    res_sign_df = all_sign_res[[name]],
    rlt, 
    coldata = phenotable,
    condition_col = "condition",
    conditions = c("AMR", "NR"), scale_rows = FALSE # scale_rows = FALSE - actual counts.
  )
}

volcano_plots$humoral
```
### Анализ обогащения гуморальное отторжение###
```{r}
up_humoral <- res_sign_humoral %>% 
  as.data.frame() %>% 
  filter(log2FoldChange > 0)
down_humoral <- res_sign_humoral %>% 
  as.data.frame() %>% 
  filter(log2FoldChange < 0)
rownames(up_humoral)
rownames(down_humoral)
```
**Конвертация в MIMATID** 
```{r}
converted_up_humoral <- miRNAVersionConvert(rownames(up_humoral))
converted_up_humoral
```
**Запрос таргетов из базы multiMiR** 
```{r}
targets_humoral_down <- unique(get_multimir(org = "hsa", mirna = converted_up_humoral$Accession, table = "validated")@data$target_symbol)
#writeLines(targets_down, "targets_down150_list.txt")
```
**Анализ обогащения из базы биологических процессов**
```{r}
#msig_go_bp <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")
# targets_down <- readLines("targets_down150_list.txt")
# targets_up <- readLines("targets_up150_list.txt")

GO_enrich_down_humoral_bp <- enrichGO(
  gene          = targets_humoral_down,  
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP", 
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05
)
```

**Визуализация**
```{r}

library(scales)

p1 <- dotplot(GO_enrich_down_humoral_bp, showCategory = 20) +
  ggtitle("GO Enrichment for DOWNregulated targets") +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.text.y = element_text(size = 20)
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 40))

p1

ggsave("./pictures_transpl/transpl_GO_enrichment_dotplot_down_humoral_bp.tiff", plot = p1, width = 16, height = 10, dpi = 300)
```
```{r}
GO_enrich_DOWN_humoral_BP <- enrichplot::pairwise_termsim(GO_enrich_down_humoral_bp, method = "JC")

plt <- emapplot(GO_enrich_DOWN_humoral_BP, 
         repel = TRUE,
         showCategory = 20) +
  ggtitle("Biological processes for DOWNregulated targets for humoral") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 3)
    )    

plt

ggsave("./pictures_transpl/transpl_GO_enrichment_emapplot_down_humoral_BP.tiff", plot = plt, width = 16, height = 10, dpi = 300)
```
```{r}
GO_enrich_down_humoral_bp@result$Description[1:20]
```

**Визуализация путей KEGG**

```{r}
msig_go_bp <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG")


GO_enrich_down_bp_KEGG <- enricher(gene = targets_humoral_down, TERM2GENE = msig_go_bp[, c("gs_name", "gene_symbol")])

plt <- dotplot(GO_enrich_down_bp_KEGG, showCategory = 20) + 
  ggtitle("GO Enrichment for DOWNregulated targets") + 
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 12)
  )

plt
ggsave("./pictures_transpl/transpl_GO_enrichment_dotplot_down_humoral_KEGG.tiff", plot = p1, width = 16, height = 10, dpi = 300)
```

**Анализ обогащения TCAD**
```{r}
res_sign_TCAD <- subset(res_TCAD, padj < 0.05 & !is.na(padj) & abs(log2FoldChange) > 1.0)

up_TCAD <- res_sign_TCAD %>% 
  as.data.frame() %>% 
  filter(log2FoldChange > 0)
down_TCAD <- res_sign_TCAD %>% 
  as.data.frame() %>% 
  filter(log2FoldChange < 0)
rownames(up_TCAD)
rownames(down_TCAD)
```



```{r}
converted_down_TCAD <- miRNAVersionConvert(rownames(down_TCAD))
converted_down_TCAD
```
**Запрос таргетов из базы multiMi**
```{r}
targets_TCAD_up <- unique(get_multimir(org = "hsa", mirna = converted_down_TCAD$Accession, table = "validated")@data$target_symbol)
```
**Анализ обогащения из базы биологических процессов**
```{r}
GO_enrich_up_TCAD_bp <- enrichGO(
  gene          = targets_TCAD_up,  
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP", 
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05
)

p1 <- dotplot(GO_enrich_up_TCAD_bp, showCategory = 20) + 
  ggtitle("GO Enrichment for UPregulated targets TCAD") + 
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.text.y = element_text(size = 18)
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 40))

p1

ggsave("./pictures_transpl/transpl_GO_enrichment_dotplot_up_TCAD_bp.tiff", plot = p1, width = 16, height = 10, dpi = 300)
```


```{r}
GO_enrich_UP_TCAD_bp <- enrichplot::pairwise_termsim(GO_enrich_up_TCAD_bp, method = "JC")

plt <- emapplot(GO_enrich_UP_TCAD_bp, 
         repel = TRUE,
         showCategory = 20) +
  ggtitle("Biological processes for UPregulated targets for TCAD") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 3)
    )    

plt

ggsave("./pictures_transpl/transpl_GO_enrichment_emapplot_UP_TCAD_BP.tiff", plot = plt, width = 16, height = 10, dpi = 300)
```
```{r}
GO_enrich_up_TCAD_bp@result$Description[1:20]
```


**Визуализация путей KEGG**
```{r}
msig_go_bp <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG")


GO_enrich_up_TCAD_KEGG <- enricher(gene = targets_TCAD_up, TERM2GENE = msig_go_bp[, c("gs_name", "gene_symbol")])

plt <- dotplot(GO_enrich_up_TCAD_KEGG, showCategory = 20) +
  ggtitle("GO Enrichment for UPregulated targets") +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 12)
  )
plt
ggsave("./pictures_transpl/transpl_GO_enrichment_emapplot_UP_TCAD_KEGG.tiff", plot = plt, width = 16, height = 10, dpi = 300)
```

	
```{r}
all_df <- read.csv("data/mapped.csv", header = TRUE, sep = ",")
all_df <- all_df[, -c(1, 2)]
cols <- c("exact.miRNA", "hairpin.miRNA", "mature.tRNA", "primary.tRNA", "snoRNA", "rRNA", "ncrna.others", "mRNA", "isomiR.miRNA")
all_df[cols] <- lapply(all_df[cols], function(x) ifelse(x == "", NA, x))
str(all_df)
```

```{r}
all_df$merged_col <- apply(all_df[, c("exact.miRNA", "hairpin.miRNA", "mature.tRNA", "primary.tRNA", "snoRNA", "rRNA", "ncrna.others", "mRNA", "isomiR.miRNA")], 1, function(x) na.omit(x)[1])
all_df <- all_df[, -c(1:9)]
all_df
```

```{r}
collapsed_df <- all_df %>%
  group_by(merged_col) %>%
  summarise(across(everything(), sum, na.rm = FALSE)) %>%
  as.data.frame()  
rownames(collapsed_df) <- collapsed_df$merged_col
collapsed_df$merged_col <- NULL 
# write.csv(collapsed_df, file="./collapsed_df.csv")
```

```{r}
# collapsed_df <- read.csv("./collapsed_df.csv", header = TRUE, sep = ",")
rownames(collapsed_df) <- collapsed_df$X
collapsed_df$X <- NULL
collapsed_df
```
```{r}
colnames(collapsed_df) <- gsub("^X", "", colnames(collapsed_df))
common_samples <- intersect(colnames(collapsed_df), coldata$sample)
collapsed_df <- rownames_to_column(collapsed_df, var = "ncRNA")

collapsed_df <- collapsed_df[, c("ncRNA", common_samples)] 
# collapsed_df <- collapsed_df[, rownames(coldata)] #ранжирую по колонке в counts так же как и названия строк в coldata
rownames(collapsed_df) <- collapsed_df$ncRNA
collapsed_df$ncRNA <- NULL
head(collapsed_df)
```
