---
title: "micro-RNA Heart Transplantation"
output:
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = TRUE,
	warning = FALSE
)
```

### Data importing ###

```{r}
source(here("config.R"))
source(here("Transplantation", "config_t.R"))
source(here("src", "data_processing.R"))
source(here("src", "visualization.R"))
source(here("src", "calculations.R"))
data <- load_and_process_data()
counts <- data$counts
phenotable <- data$phenotable
summed_anno <- load_sample_annotation(phenotable = data$phenotable)
```
**Venn plot (>150 counts)**

```{r}
venn_plot <- create_venn_diagram(
  counts = data$counts,
  phenotable = data$phenotable
)
venn_plot$plot
```
**Unique miRs**
```{r}
get_unique_venn_elements(venn_plot$venn_list)
```
**Barplots**
```{r}
create_barplot_all_dataset(data = summed_anno)
```
```{r}
create_barplot_all_dataset_normalized(data = summed_anno)
```
```{r}
grouped_barplots <- create_grouped_barplots(data = summed_anno, phenotable = data$phenotable)
grouped_barplots$counts
#grouped_barplots$proportion
```
###  Run Differential Expression Analysis ###
```{r}
dds <- create_dds(counts, coldata = phenotable, ref_level = "NR", condition_col = "condition", check_rank = TRUE)
dds <- DESeq(dds, fitType = "parametric")
dds 
```
**transformation**

• The plot should show a nearly horizontal line, meaning the SD is independent of the mean expression.
• This means the transformation has successfully stabilized the variance, making the data more suitable for further analysis.
```{r}
rlt_list <- auto_transform_dds(dds)
rlt <- rlt_list$transformed
rlt_list
```
**PCA plot**
```{r}
create_pca_plot(rlt, condition_col = "condition")
```

**Plot a heatmap of most variable genes**
```{r}
create_heatmap_most_expressed(rlt, top_n = 30, group_col = "condition", output_file = NULL)
```
**Plot of the distance between samples heatmap**

```{r}
create_sample_distance_heatmap(rlt, group_col = "condition", output_file = NULL)
```
## Results ##
```{r}
all_res <- list()
all_sign_res <- list()
for (name in names(CONTRASTS)) {
  res <- get_significant_results(
    dds = dds,
    contrast = CONTRASTS[[name]],
    output_res = file.path("results", paste0("results_", name, ".csv")),
    output_sign_res = file.path("results", paste0("sign_results_", name, ".csv")),
    alpha = 0.05, lfc_threshold = 1)
  all_res[[name]] <- res$res_df         
  all_sign_res[[name]] <- res$res_sign_df
}

all_sign_res$humoral
all_sign_res$cellular
all_sign_res$CAV
```
**MA plot**
```{r}
ma_plots <- list()
for(name in names(all_res)) {
  ma_plots[[name]] <- plot_ma_from_res(
    res_df = all_res[[name]], 
    alpha = 0.05, 
    contrast_name = name,
    output_file = file.path("figures", paste0("ma_plot_", name, ".png"))
  )
}

ma_plots$humoral
ma_plots$cellular
ma_plots$CAV
```
**Volcano plot**

```{r}
volcano_plots <- list()
for(name in names(all_res)) {
  volcano_plots[[name]] <- create_volcano_plot(
    res_df = all_res[[name]],
    res_sign_df = all_sign_res[[name]],
    contrast_name = name,
    output_file = file.path("figures", paste0("volcano_", name, ".png"))
  )
}

volcano_plots$humoral
volcano_plots$cellular
volcano_plots$CAV
```

**Create boxplot for specific gene**

```{r}
create_gene_boxplot(dds, "hsa-miR-100-5p", output_file = NULL)
```
**Plot a heatmap of diff expressed genes**
```{r}
heatmap_plots <- list()
for(name in names(CONTRASTS)) {
  heatmap_plots[[name]] <- create_heatmap_diff_expressed(
    res_sign_df = all_sign_res[[name]],
    rlt = rlt,
    coldata = phenotable,
    conditions = CONTRASTS[[name]][2:3],
    output_file = file.path("results", paste0("heatmap_diff_", name, ".png"))
  )
}

heatmap_plots$humoral
heatmap_plots$cellular
heatmap_plots$CAV
```
### Enrichment analysis ###
**Searching for targets**
```{r}
source(here("src", "enrichment.R"))
targets_list <- list()
for(name in names(all_sign_res)) {
  targets_list[[name]] <- convert_miRNA_versions(
    res_sign_df = all_sign_res[[name]],
    contrast_name = name,
    table = "validated"
  )
}
```
**Perform KEGG enrichment analysis**
```{r}
kegg_plots <- list()

for (contrast in names(targets_list)) {
  kegg_plots[[contrast]] <- list()
  
  for (direction in c("up", "down")) {
    gene_list <- targets_list[[contrast]][[direction]]
    
    if (length(gene_list) == 0) {
      message(sprintf("No genes for %s %s. Skipping...", contrast, direction))
      kegg_plots[[contrast]][[direction]] <- NULL
      next
    }
    
    kegg_plots[[contrast]][[direction]] <- perform_kegg_enrichment(
      gene_list,
      contrast_name = contrast,
      direction = direction
    )
  }
}

# пример обращения
kegg_plots$humoral$up
kegg_plots$humoral$down
kegg_plots$CAV$up
```